A Geometry for Network Visualization in \MakeLowercase{\texttt{ggplot2}}

### 2015-10-11 inst/CC/writeup.tex
finished side by side examples - problems with ggnet2 in football example

\documentclass[11pt]{article}
\usepackage{asa}
\usepackage{xspace}
\usepackage{natbib}
\usepackage{url}
\usepackage{afterpage}
\usepackage{pdflscape}
\usepackage{geometry} % see geometry.pdf on how to lay out the page. There's lots.
\geometry{letterpaper} % or letter or a5paper or ... etc
% \geometry{landscape} % rotated page geometry
% See the ``Article customise'' template for come common customisations

@@

\title{A Geometry for Network Visualization in \MakeLowercase{\texttt{ggplot2}}}
\author{Sam Tyner and Heike Hofmann}
\date{} % delete this line to display the current date
%%% BEGIN DOCUMENT
\begin{document}

\maketitle
\begin{abstract}
There are many implementations of static network visualization in R, but none are equipped with the flexibility and functionality of \texttt{ggplot2}.  \texttt{geom\_net} was created to fill this gap. Using two data frames to describe vertex and edge information, \texttt{geom\_net} makes use of the underlying structure of \texttt{ggplot2} to visualize networks. This makes it possible to easily facet networks according to covariates  or change aesthetics such as shape, size, or color according to additional edge or vertex information.
\end{abstract}
<<chunk_setup, echo=FALSE>>=
library(knitr)
opts_chunk$set(tidy = FALSE, echo=FALSE, cache=FALSE, eval = TRUE, message = FALSE, warning = FALSE,  fig.height = 8, fig.width = 8, fig.align = 'center',fig.show='hide')
@
<<readingdata, eval=TRUE, echo=FALSE>>=
library(ggplot2) # needs to be the version with geom_net!
 theme_net <- theme(aspect.ratio = 1, line = element_blank(), panel.background = element_blank(), plot.background = element_blank(), axis.text=element_blank(),axis.title=element_blank())
 data(blood, football, theme_elements, lesmis,madmen, mm_directed,email)
names(blood$vertices)[1] <- 'label'
#load("data/madmen.rda") # relative path should work for both of us
#load("data/mm_directed.rda")
#names(madmen$vertices)[1] <- 'label'
names(theme_elements$vertices)[1] <- 'label'
football$vertices$label <- as.factor(football$vertices$label)
football$vertices$value <- as.factor(football$vertices$value)
football$edges$in.conf <- 0
for (i in 1:nrow(football$edges)){
  idx1 <- which(football$vertices$label == football$edges$from[i])
  idx2 <- which(football$vertices$label == football$edges$to[i])
  if (football$vertices$value[idx1] == football$vertices$value[idx2]){
    football$edges$in.conf[i] <- 1
  }
}
#load("data/email.rda")
#email.net <- read.csv("data/email-data.csv")
#email.names <- data.frame(label = as.factor(unique(c(as.character(email.net$From),as.character(email.net$to)))))
#library(plyr)
#email.net <- ddply(email.net, .(eID), transform, nrecipients=length(to))
#employee <- read.csv("data/EmployeeRecords.csv")
#employee <- merge(email.names, employee, by.x="label", by.y="EmailAddress", all.x=T)
email.net <- email$edges
employee <- email$nodes
library(RColorBrewer)
cols <- brewer.pal(8, name="Set1")
load("data/protein.rda")

#yeast <- list(edges = protein, vertices = data.frame(label = min(yeast):max(yeast)))
yeast <- protein
@

@@

This network is shown in figure~\ref{fig.cap:madmen_ex}.
<<madmen_ex, dependson='madmen-prep', fig.cap = 'Network of Relationships in Mad Men'>>=
ggplot(data = madmen$edges, aes(from_id = Name1, to_id = Name2)) +
  geom_net(vertices = madmen$vertices, vsize=3, vlabel= TRUE, ecolour="grey30",
           aes(vcolour=c( "#FF69B4", "#0099ff")[as.numeric(Gender)])) +
  expand_limits(x = c(0,1.25), y = c(0,1)) + theme_net
@

@@

\section{Usage Examples}
In this section, we demonstrate the current capabilities of \texttt{geom\_net} in a series of diverse examples.
  \subsection{Blood Donation}
  In this directed network, there are eight vertices and 27 edges.  The vertices represent the eight different blood types in humans that are most important for donation: the ABO blood types A, B, AB, and O, combined with the RhD positive (+) and negative (-) types. The edges are directed: a person whose blood type is that of a \emph{from} vertex can to donate blood to a person whose blood type is that of a corresponding \emph{to} vertex. In the example below, loops are removed because loops exist on every vertex in this example, as blood between two people of matching ABO and RhD type can always be exchanged.
<<blood_ex, echo=TRUE, fig.cap= "Directed network of blood type donations in humans.">>=
ggplot(data = blood$edges, aes(from_id = from, to_id = to)) +
  geom_net(vertices = blood$vertices, vcolour = I('red'), layout = 'circle',
           vlabel = TRUE, vsize = I(3), directed = TRUE) +
  expand_limits(x = c(0,1), y = c(0,1)) + theme_net
@
This network is shown in figure \ref{fig.cap:blood_ex}.  Here, we have used the aesthetics \texttt{vcolour} and \texttt{vsize} set to identity values to change the size and color of all vertices. We have also used the \texttt{layout} and \texttt{vlabel} arguments to change the default layout to a circle layout and to print the blood types, respectively. The circle layout places blood types of the same ABO type next to each other and spreads the vertices out far enough to distinguish between the various ``in" and ``out" types.  You can tell clearly from this plot that the O- type is the universal donor: it has out degree of seven and in degree of zero. Additionally, you can see that the AB+ type is the universal recepient, with in degree of seven and out degree of zero. Anyone looking at this plot can quickly determine which type(s) of blood they can receive and which type(s) can receive their blood.
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figure/blood_ex-1.pdf}
\caption{\label{fig.cap:blood_ex} Network of blood donation possibilities in humans by ABO and RhD blood types.}
\end{figure}
\afterpage{\clearpage}

\subsection{Email Network}
  This email network comes from the 2014 VAST Challenge \citep{emailnet}. It is a directed network of emails between company employees with 55 vertices and 9,063 edges. Each vertex is an employee of the company, and each edge is an email sent from one employee to one or more other employees. The arrow of the directed edge points to the recipient(s) of the email. The network contains two business weeks of emails across the entire company. In order to better visualize the structure of the communication network between employees, all emails that were sent out to all employees are removed in the subsequent examples.
<<email_ex, fig.cap = 'All emails within a company over a two week period.',echo=TRUE>>=
ggplot(data = subset(email.net, nrecipients < 54), aes(from_id = From, to_id = to)) +
  geom_net(vertices = employee,ealpha = 0.1, vsize = I(4), directed = TRUE,
           aes(vcolour=cols[as.numeric(CurrentEmploymentType)])) +
  expand_limits(x=c(0,1), y = c(0,1)) + theme_net
@
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figure/email_ex-1.pdf}
\caption{\label{fig.cap:email_ex} Email network within a company over a two week period.}
\end{figure}
\afterpage{\clearpage}

@@

<<email_ex_facet, fig.height=8, fig.width=8, fig.cap = 'All emails within a company over a two week period facetted by day of the month.',echo=TRUE,eval=FALSE>>=
ggplot(data = subset(email.net, nrecipients < 54), aes(from_id = From, to_id = to)) +
  geom_net(vertices = employee, ealpha = 0.2, vsize = I(2), directed = TRUE,
           aes(vcolour=cols[as.numeric(CurrentEmploymentType)])) +
  expand_limits(x=c(0,1), y = c(0,1)) + facet_wrap(~day, nrow = 2) + theme_net
@

@@

<<theme_ex, echo=TRUE, fig.cap= 'Inheritance structure for \\texttt{theme} elements in \\texttt{ggplot2}.'>>=
ggplot(data = theme_elements$edges, aes(from_id = parent, to_id = child)) +
  geom_net(vertices = theme_elements$vertices, directed = TRUE, vlabel = TRUE) +
  expand_limits(x = c(0,1.2), y = c(0,1)) +
  theme_net
@
\includegraphics[width=\textwidth]{figure/theme_ex-1.pdf}
\caption{\label{fig.cap:theme_ex} Inheritance structure of \texttt{ggplot2} theme elements. This is a recreation of the graph found at \protect\url{http://docs.ggplot2.org/current/theme.html}.}
\afterpage{\clearpage}

The inheritance structure is plotted in figure \ref{fig.cap:theme_ex}.  In this plot, it is easy to quickly determine the parent and child vertices. Using this plot made creation of the \texttt{theme\_net} object used throughout these examples very simple. We just made set each of the major parent elements, \texttt{text}, \texttt{rect}, and \texttt{line} to \texttt{element\_blank()} and then set the aspect ratio equal to one.
  \subsection{Mad Men Networks}
  The following code creates the network example given in the introduction. We changed the vertex size and edge color for all vertices and edges, included vertex labels, and colored the vertices according to the character's gender.
<<print_ex, ref.label="madmen_ex", eval=FALSE, echo=TRUE,cache=FALSE>>=
@
 There is another Mad Men network included in the \texttt{gcookbook} package \citep{madmen}. It is a directed network, also of romantic relationships between characters, but it also includes advances made by one character that were rejected by the other.  For example, Roger Sterling made advances toward Betty Draper, but Betty refused him, and so there is a directed edge going from Roger to Betty, but not from Betty to Roger.  If the advance is reciprocated, like between Sal Romano and the Bellhop, there are two directed edges between the two vertices.
<<madmen2_ex,echo=TRUE, fig.cap= 'A directed network of relationships in Mad Men.'>>=
ggplot(data = mm.directed$edges, aes(from_id = Name1, to_id = Name2)) +
  geom_net(vertices = mm.directed$vertices, directed = T, vlabel = T,
           vsize = I(2.5), layout = 'fruchtermanreingold',
           aes(vcolour=c( "#FF69B4", "#0099ff")[as.numeric(Gender)])) +
  expand_limits(x = c(0,1.1), y = c(0,1)) + theme_net
@
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figure/madmen2_ex-1.pdf}
\caption{\label{fig.cap:madmen2_ex} A directed network of relationships in Mad Men.}
\end{figure}
This network is shown in figure \ref{fig.cap:madmen2_ex}.  This network is much more connected than the previous Mad Men example.  It also allows us to see much more of the drama in the show. For instance, you can see that Roger made advances towards Betty, his business partner's wife, which is much more scandalous than what we can see in the first network.

  \subsection{College Football}
  This next example comes from M.E.J. Newman's network data web page \citep{football}.  It is an undirected network consisting of all regular season college football games played between Division I schools in Fall of 2000.  There are 115 vertices and 613 edges: each vertex represents a school, and an edge represents a game played between two schools. There is an additional variable in the vertex data frame corresponding to the conference each team belongs to, and there is an additional variable in the edge data frame that is equal to one if the game occured between teams in the same conference or zero if the game occured between teams in different conferences.
<<football_ex, echo=TRUE, fig.cap="All regular season Division I college football games in the Fall 2000 season.">>=
colors <- brewer.pal(12, name = 'Paired')
ggplot(data = football$edges, aes(from_id = from, to_id = to,
                                  elinetype = in.conf + 1)) +
  geom_net(vertices = football$vertices, ealpha = 0.3, vsize = 2, vlabel = TRUE,
           aes(vcolour=colors[as.numeric(factor(value))])) +
  expand_limits(x = c(0,1), y = c(0,1)) + theme_net
@
The network of football games is given in figure \ref{fig.cap:football_ex}. Here, we have changed the \texttt{elinetype} aesthetic to correspond to games that occur between teams in the same conference or different conferences.  These lines are dotted and solid, respectively. We have also assigned a different color to each conference, and the vertices and their labels are colored according to their conference. This coloring and changing of the linetypes make the structure of the game network easier to view.  There is one conference consisting of Navy, Notre Dame, Utah State, Central Florida, and Connecticut, which is spread out all over the network, whereas most other conferences' teams are all very close to each other because they play within conference much more than they play out of conference.  At the time, these five schools were all independents and did not have a home conference.  Without the coloring capability, we would not have been able to pick out that difference as easily.
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figure/football_ex-1.pdf}
\caption{\label{fig.cap:football_ex} The network of regular season Division I college football games in the season of fall 2000. The vertices and their labels are colored by conference.}
\end{figure}
\afterpage{\clearpage}

  \subsection{Les Mis\'{e}rables}
  This next network comes from \citet{lesmis}. It is an undirected network of coappearances of characters in Victor Hugo's \emph{Les Mis\'{e}rables}.  There are 77 vertices representing each of the 77 characters in the book, and an edge connects two vertices if those two characters appear in the same chapter of the book.  There are 254 edges in this network. The edges are also weighted by the number of coappearances. The largest weighting is 31, between the characters Jean Valjean and Cosette.  This network is shown in figure \ref{fig.cap:lesmis_ex}. This network reveals many important character groups, as well as the main characters in the story.  The main characters are the ones with high degree, like Valjean, Cosette, and Javert.  There is an extremely wide edge connecting Valjean and Cosette, which implies that they are strongly connected; in fact, Cosette is Valjean's adopted daughter.
<<lesmis_ex, echo=TRUE, fig.cap="Coappearance network of characters in Victor Hugo's \\emph{Les Mis\\'{e}rables}">>=
ggplot(data = lesmis$edges, aes(from_id = from, to_id = to,
                                esize = value/mean(value))) +
  geom_net(vertices = lesmis$vertices, ealpha = 7/10,
           vcolour = I("red"), vsize = I(2.5),
           layout = 'fruchtermanreingold', vlabel=TRUE) +
  expand_limits(x = c(0,1.1), y = c(0,1)) + theme_net
@
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figure/lesmis_ex-1.pdf}
\caption{\label{fig.cap:lesmis_ex} Co-appearance network of characters in Victor Hugo's \emph{Les Mis\'{e}rables}.}
\end{figure}
\afterpage{\clearpage}

>>>>

\documentclass[11pt]{article}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}
\usepackage{asa}
\usepackage{xspace}
\usepackage{natbib}
\usepackage{url}
\usepackage{afterpage}
\usepackage{pdflscape}
\usepackage{geometry} % see geometry.pdf on how to lay out the page. There's lots.
\geometry{letterpaper} % or letter or a5paper or ... etc
% \geometry{landscape} % rotated page geometry
% See the ``Article customise'' template for come common customisations

@@

\title{A Geometry for Network Visualization in \MakeLowercase{\texttt{ggplot2}}}
\author{Sam Tyner and Heike Hofmann}
\date{} % delete this line to display the current date
%%% BEGIN DOCUMENT
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

\maketitle
\begin{abstract}
There are many implementations of static network visualization in R, but none are equipped with the flexibility and functionality of \texttt{ggplot2}.  \texttt{geom\_net} was created to fill this gap. Using two data frames to describe vertex and edge information, \texttt{geom\_net} makes use of the underlying structure of \texttt{ggplot2} to visualize networks. This makes it possible to easily facet networks according to covariates  or change aesthetics such as shape, size, or color according to additional edge or vertex information.
\end{abstract}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): object 'email' not found}}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): object 'email' not found}}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in readChar(con, 5L, useBytes = TRUE): cannot open the connection}}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): object 'protein' not found}}\end{kframe}
\end{knitrout}

@@

This network is shown in figure~\ref{fig.cap:madmen_ex}.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): could not find function "{}geom\_net"{}}}\end{kframe}
\end{knitrout}

@@

\section{Usage Examples}
In this section, we demonstrate the current capabilities of \texttt{geom\_net} in a series of diverse examples.
  \subsection{Blood Donation}
  In this directed network, there are eight vertices and 27 edges.  The vertices represent the eight different blood types in humans that are most important for donation: the ABO blood types A, B, AB, and O, combined with the RhD positive (+) and negative (-) types. The edges are directed: a person whose blood type is that of a \emph{from} vertex can to donate blood to a person whose blood type is that of a corresponding \emph{to} vertex. In the example below, loops are removed because loops exist on every vertex in this example, as blood between two people of matching ABO and RhD type can always be exchanged.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{ggplot}\hlstd{(}\hlkwc{data} \hlstd{= blood}\hlopt{$}\hlstd{edges,} \hlkwd{aes}\hlstd{(}\hlkwc{from_id} \hlstd{= from,} \hlkwc{to_id} \hlstd{= to))} \hlopt{+}
  \hlkwd{geom_net}\hlstd{(}\hlkwc{vertices} \hlstd{= blood}\hlopt{$}\hlstd{vertices,} \hlkwc{vcolour} \hlstd{=} \hlkwd{I}\hlstd{(}\hlstr{'red'}\hlstd{),} \hlkwc{layout} \hlstd{=} \hlstr{'circle'}\hlstd{,}
           \hlkwc{vlabel} \hlstd{=} \hlnum{TRUE}\hlstd{,} \hlkwc{vsize} \hlstd{=} \hlkwd{I}\hlstd{(}\hlnum{3}\hlstd{),} \hlkwc{directed} \hlstd{=} \hlnum{TRUE}\hlstd{)} \hlopt{+}
  \hlkwd{expand_limits}\hlstd{(}\hlkwc{x} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,}\hlnum{1}\hlstd{),} \hlkwc{y} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,}\hlnum{1}\hlstd{))} \hlopt{+} \hlstd{theme_net}
\end{alltt}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): could not find function "{}geom\_net"{}}}\end{kframe}
\end{knitrout}
This network is shown in figure \ref{fig.cap:blood_ex}.  Here, we have used the aesthetics \texttt{vcolour} and \texttt{vsize} set to identity values to change the size and color of all vertices. We have also used the \texttt{layout} and \texttt{vlabel} arguments to change the default layout to a circle layout and to print the blood types, respectively. The circle layout places blood types of the same ABO type next to each other and spreads the vertices out far enough to distinguish between the various ``in" and ``out" types.  You can tell clearly from this plot that the O- type is the universal donor: it has out degree of seven and in degree of zero. Additionally, you can see that the AB+ type is the universal recepient, with in degree of seven and out degree of zero. Anyone looking at this plot can quickly determine which type(s) of blood they can receive and which type(s) can receive their blood.
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figure/blood_ex-1.pdf}
\caption{\label{fig.cap:blood_ex} Network of blood donation possibilities in humans by ABO and RhD blood types.}
\end{figure}
\afterpage{\clearpage}

\subsection{Email Network}
  This email network comes from the 2014 VAST Challenge \citep{emailnet}. It is a directed network of emails between company employees with 55 vertices and 9,063 edges. Each vertex is an employee of the company, and each edge is an email sent from one employee to one or more other employees. The arrow of the directed edge points to the recipient(s) of the email. The network contains two business weeks of emails across the entire company. In order to better visualize the structure of the communication network between employees, all emails that were sent out to all employees are removed in the subsequent examples.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{ggplot}\hlstd{(}\hlkwc{data} \hlstd{=} \hlkwd{subset}\hlstd{(email.net, nrecipients} \hlopt{<} \hlnum{54}\hlstd{),} \hlkwd{aes}\hlstd{(}\hlkwc{from_id} \hlstd{= From,} \hlkwc{to_id} \hlstd{= to))} \hlopt{+}
  \hlkwd{geom_net}\hlstd{(}\hlkwc{vertices} \hlstd{= employee,}\hlkwc{ealpha} \hlstd{=} \hlnum{0.1}\hlstd{,} \hlkwc{vsize} \hlstd{=} \hlkwd{I}\hlstd{(}\hlnum{4}\hlstd{),} \hlkwc{directed} \hlstd{=} \hlnum{TRUE}\hlstd{,}
           \hlkwd{aes}\hlstd{(}\hlkwc{vcolour}\hlstd{=cols[}\hlkwd{as.numeric}\hlstd{(CurrentEmploymentType)]))} \hlopt{+}
  \hlkwd{expand_limits}\hlstd{(}\hlkwc{x}\hlstd{=}\hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,}\hlnum{1}\hlstd{),} \hlkwc{y} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,}\hlnum{1}\hlstd{))} \hlopt{+} \hlstd{theme_net}
\end{alltt}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in subset(email.net, nrecipients < 54): object 'email.net' not found}}\end{kframe}
\end{knitrout}
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figure/email_ex-1.pdf}
\caption{\label{fig.cap:email_ex} Email network within a company over a two week period.}
\end{figure}
\afterpage{\clearpage}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{ggplot}\hlstd{(}\hlkwc{data} \hlstd{=} \hlkwd{subset}\hlstd{(email.net, nrecipients} \hlopt{<} \hlnum{54}\hlstd{),} \hlkwd{aes}\hlstd{(}\hlkwc{from_id} \hlstd{= From,} \hlkwc{to_id} \hlstd{= to))} \hlopt{+}
  \hlkwd{geom_net}\hlstd{(}\hlkwc{vertices} \hlstd{= employee,} \hlkwc{ealpha} \hlstd{=} \hlnum{0.2}\hlstd{,} \hlkwc{vsize} \hlstd{=} \hlkwd{I}\hlstd{(}\hlnum{2}\hlstd{),} \hlkwc{directed} \hlstd{=} \hlnum{TRUE}\hlstd{,}
           \hlkwd{aes}\hlstd{(}\hlkwc{vcolour}\hlstd{=cols[}\hlkwd{as.numeric}\hlstd{(CurrentEmploymentType)]))} \hlopt{+}
  \hlkwd{expand_limits}\hlstd{(}\hlkwc{x}\hlstd{=}\hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,}\hlnum{1}\hlstd{),} \hlkwc{y} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,}\hlnum{1}\hlstd{))} \hlopt{+} \hlkwd{facet_wrap}\hlstd{(}\hlopt{~}\hlstd{day,} \hlkwc{nrow} \hlstd{=} \hlnum{2}\hlstd{)} \hlopt{+} \hlstd{theme_net}
\end{alltt}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{ggplot}\hlstd{(}\hlkwc{data} \hlstd{= theme_elements}\hlopt{$}\hlstd{edges,} \hlkwd{aes}\hlstd{(}\hlkwc{from_id} \hlstd{= parent,} \hlkwc{to_id} \hlstd{= child))} \hlopt{+}
  \hlkwd{geom_net}\hlstd{(}\hlkwc{vertices} \hlstd{= theme_elements}\hlopt{$}\hlstd{vertices,} \hlkwc{directed} \hlstd{=} \hlnum{TRUE}\hlstd{,} \hlkwc{vlabel} \hlstd{=} \hlnum{TRUE}\hlstd{)} \hlopt{+}
  \hlkwd{expand_limits}\hlstd{(}\hlkwc{x} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,}\hlnum{1.2}\hlstd{),} \hlkwc{y} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,}\hlnum{1}\hlstd{))} \hlopt{+}
  \hlstd{theme_net}
\end{alltt}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): could not find function "{}geom\_net"{}}}\end{kframe}
\end{knitrout}
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figure/theme_ex-1.pdf}
\caption{\label{fig.cap:theme_ex} Inheritance structure of \texttt{ggplot2} theme elements. This is a recreation of the graph found at \protect\url{http://docs.ggplot2.org/current/theme.html}.}
\end{figure}
\afterpage{\clearpage}

The inheritance structure is plotted in figure \ref{fig.cap:theme_ex}.  In this plot, it is easy to quickly determine the parent and child vertices. Using this plot made creation of the \texttt{theme\_net} object used throughout these examples very simple. We just made set each of the major parent elements, \texttt{text}, \texttt{rect}, and \texttt{line} to \texttt{element\_blank()} and then set the aspect ratio equal to one.
  \subsection{Mad Men Networks}
  The following code creates the network example given in the introduction. We changed the vertex size and edge color for all vertices and edges, included vertex labels, and colored the vertices according to the character's gender.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{ggplot}\hlstd{(}\hlkwc{data} \hlstd{= madmen}\hlopt{$}\hlstd{edges,} \hlkwd{aes}\hlstd{(}\hlkwc{from_id} \hlstd{= Name1,} \hlkwc{to_id} \hlstd{= Name2))} \hlopt{+}
  \hlkwd{geom_net}\hlstd{(}\hlkwc{vertices} \hlstd{= madmen}\hlopt{$}\hlstd{vertices,} \hlkwc{vsize}\hlstd{=}\hlnum{3}\hlstd{,} \hlkwc{vlabel}\hlstd{=} \hlnum{TRUE}\hlstd{,} \hlkwc{ecolour}\hlstd{=}\hlstr{"grey30"}\hlstd{,}
           \hlkwd{aes}\hlstd{(}\hlkwc{vcolour}\hlstd{=}\hlkwd{c}\hlstd{(} \hlstr{"#FF69B4"}\hlstd{,} \hlstr{"#0099ff"}\hlstd{)[}\hlkwd{as.numeric}\hlstd{(Gender)]))} \hlopt{+}
  \hlkwd{expand_limits}\hlstd{(}\hlkwc{x} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,}\hlnum{1.25}\hlstd{),} \hlkwc{y} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,}\hlnum{1}\hlstd{))} \hlopt{+} \hlstd{theme_net}
\end{alltt}
\end{kframe}
\end{knitrout}
 There is another Mad Men network included in the \texttt{gcookbook} package \citep{madmen}. It is a directed network, also of romantic relationships between characters, but it also includes advances made by one character that were rejected by the other.  For example, Roger Sterling made advances toward Betty Draper, but Betty refused him, and so there is a directed edge going from Roger to Betty, but not from Betty to Roger.  If the advance is reciprocated, like between Sal Romano and the Bellhop, there are two directed edges between the two vertices.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{ggplot}\hlstd{(}\hlkwc{data} \hlstd{= mm.directed}\hlopt{$}\hlstd{edges,} \hlkwd{aes}\hlstd{(}\hlkwc{from_id} \hlstd{= Name1,} \hlkwc{to_id} \hlstd{= Name2))} \hlopt{+}
  \hlkwd{geom_net}\hlstd{(}\hlkwc{vertices} \hlstd{= mm.directed}\hlopt{$}\hlstd{vertices,} \hlkwc{directed} \hlstd{= T,} \hlkwc{vlabel} \hlstd{= T,}
           \hlkwc{vsize} \hlstd{=} \hlkwd{I}\hlstd{(}\hlnum{2.5}\hlstd{),} \hlkwc{layout} \hlstd{=} \hlstr{'fruchtermanreingold'}\hlstd{,}
           \hlkwd{aes}\hlstd{(}\hlkwc{vcolour}\hlstd{=}\hlkwd{c}\hlstd{(} \hlstr{"#FF69B4"}\hlstd{,} \hlstr{"#0099ff"}\hlstd{)[}\hlkwd{as.numeric}\hlstd{(Gender)]))} \hlopt{+}
  \hlkwd{expand_limits}\hlstd{(}\hlkwc{x} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,}\hlnum{1.1}\hlstd{),} \hlkwc{y} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,}\hlnum{1}\hlstd{))} \hlopt{+} \hlstd{theme_net}
\end{alltt}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in ggplot(data = mm.directed\$edges, aes(from\_id = Name1, to\_id = Name2)): object 'mm.directed' not found}}\end{kframe}
\end{knitrout}
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figure/madmen2_ex-1.pdf}
\caption{\label{fig.cap:madmen2_ex} A directed network of relationships in Mad Men.}
\end{figure}
This network is shown in figure \ref{fig.cap:madmen2_ex}.  This network is much more connected than the previous Mad Men example.  It also allows us to see much more of the drama in the show. For instance, you can see that Roger made advances towards Betty, his business partner's wife, which is much more scandalous than what we can see in the first network.

  \subsection{College Football}
  This next example comes from M.E.J. Newman's network data web page \citep{football}.  It is an undirected network consisting of all regular season college football games played between Division I schools in Fall of 2000.  There are 115 vertices and 613 edges: each vertex represents a school, and an edge represents a game played between two schools. There is an additional variable in the vertex data frame corresponding to the conference each team belongs to, and there is an additional variable in the edge data frame that is equal to one if the game occured between teams in the same conference or zero if the game occured between teams in different conferences.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{colors} \hlkwb{<-} \hlkwd{brewer.pal}\hlstd{(}\hlnum{12}\hlstd{,} \hlkwc{name} \hlstd{=} \hlstr{'Paired'}\hlstd{)}
\hlkwd{ggplot}\hlstd{(}\hlkwc{data} \hlstd{= football}\hlopt{$}\hlstd{edges,} \hlkwd{aes}\hlstd{(}\hlkwc{from_id} \hlstd{= from,} \hlkwc{to_id} \hlstd{= to,}
                                  \hlkwc{elinetype} \hlstd{= in.conf} \hlopt{+} \hlnum{1}\hlstd{))} \hlopt{+}
  \hlkwd{geom_net}\hlstd{(}\hlkwc{vertices} \hlstd{= football}\hlopt{$}\hlstd{vertices,} \hlkwc{ealpha} \hlstd{=} \hlnum{0.3}\hlstd{,} \hlkwc{vsize} \hlstd{=} \hlnum{2}\hlstd{,} \hlkwc{vlabel} \hlstd{=} \hlnum{TRUE}\hlstd{,}
           \hlkwd{aes}\hlstd{(}\hlkwc{vcolour}\hlstd{=colors[}\hlkwd{as.numeric}\hlstd{(}\hlkwd{factor}\hlstd{(value))]))} \hlopt{+}
  \hlkwd{expand_limits}\hlstd{(}\hlkwc{x} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,}\hlnum{1}\hlstd{),} \hlkwc{y} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,}\hlnum{1}\hlstd{))} \hlopt{+} \hlstd{theme_net}
\end{alltt}
